---
title: "Practical Machine Learning Project"
author: "Paaronen"
date: "27 September 2015"
output: html_document
---

## Introduction

## Loading data


```{r}
# The training data for this project are available here:
trainingUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
# read file
training <- read.csv(url(trainingUrl), na.strings=c("NA","#DIV/0!",""))

# The test data are available here:
testingUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
# read file
testing <- read.csv(url(testingUrl), na.strings=c("NA","#DIV/0!",""))

```


# Preparing datasets

I use caret-package exclusively in this assignment which can interfaces to
 194 models). To mention other useful packages there is at least  rattle and rapidminer. 


Imputation was not necessary.
In addition also first five variables which are related to observation identification and time stamping are removed since they have no meaning in prediction process. 


```{r, echo=TRUE}
library(caret)
# remove variables with nearly zero variance
nzv <- nearZeroVar(training)
training1 <- training[, -nzv]

# remove variables where the missingness is over 50% 
halfNAs <- sapply(training1, function(x) mean(is.na(x))) > 0.50
training2 <- training1[, halfNAs==F]

# Remove first five variables
modellingdata <- training2[, -(1:5)]
remove(training1)
remove(training2)

```






# Splitting data
Function createDataPartition splits data within groups 
```{r}
trainIndex <- createDataPartition(y=modellingdata$classe, p=0.6, list=FALSE)
trainData = modellingdata[trainIndex,]
testData = modellingdata[-trainIndex,]
dim(trainData); dim(testData)
set.seed(3433) # Without this you cant get the result

```






With qualitative response variable I have to leave regression models out
and focus more appropriate modeling techniques. From these I chose to try 
tree approaches:lda, rpart and fandom forest

Linear discriminant analysis applies Bayes theorem to classifying observations into the most probable class. The model is quite similar to more familiar logistic regression. Rpart is implemention to apply Classification and Regression Trees developed mostly by Breiman et al 1984. 

In random forest approach number of decision trees are constructed on bootstrapped training samples. Approach is randon in a sense that in splitting a random sample of predictors is chosen as candidates to further splitting from the full set of variables. Main note is that with this approach majority of variables are always excluded from the splitting procedure.   

# Caret has as a default resampling scheme the bootstrap.
```{r}
## linear discriminant analysis

lda.fit <- train(trainData$classe ~ ., data=trainData, method="lda")
print(lda.fit$finalModel)


# classification tree
rpart.fit <- train(trainData$classe ~ ., data=trainData, method="rpart")
print(rpart.fit$finalModel)

# random forest
fitControl <- trainControl(method="cv", number=3, verboseIter=F)
rf.fit <- train(trainData$classe ~ ., data=trainData, method="rf", trControl=fitControl)
print(rf.fit$finalModel)

```



## Comparing models

```{r}
ConfMat.lda <- confusionMatrix(testData$classe, predict(lda.fit, newdata=testData))
print(ConfMat.lda)
ConfMat.lda$overall[1]

ConfMat.rpart <- confusionMatrix(testData$classe, predict(rpart.fit, newdata=testData))
print(ConfMat.rpart)
ConfMat.rpart$overall[1]

ConfMat.rf <- confusionMatrix(testData$classe, predict(rf.fit, newdata=testData))
print(ConfMat.rf)
ConfMat.rf$overall[1]


```


# Final model with whole training data

```{r}
final.fit <- train(modellingdata$classe ~ ., data=modellingdata, method="rf", trControl=fitControl)
print(final.fit$finalModel)

```


# Submission to Coursera

```{r}

# Predict on the test data
preds <- as.character(predict(final.fit, testing))

# create function to write predictions to files
pml_write_files = function(x){
        n <- length(x)
        for(i in 1:n) {
                filename <- paste0("problem_id_", i, ".txt")
                write.table(x[i], file=filename, quote=F, row.names=F, col.names=F)
        }
} 



#
# create prediction files to submit
pml_write_files(preds)

```


